---
title: "Final Example"
author: "S. Holmes"
date: "2025-07-21"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r, setup, include = FALSE, echo = FALSE}
library(ggplot2)
library(readr)
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE)


# The above options name the code chunk 'global options.' include = FALSE means this code chunk will not be rendered in the final document. The knitr package will create the report once rendered.


# By default, the working directory is the location where the .Rmd file is saved. setwd() does not work in RMarkdown!
```

```{r, include = FALSE}
getwd()
```

# R Coding Final Title

## Student Name

### GA Tech, Summer 2025

#### Research Topic or Abstract

These data were obtained from Toxic Release Inventory (TRI) data from 1987-2022 that examined the effect of pollution on on infant birth outcomes (specifically, *birth weights*) in Louisiana's infamous **Cancer Alley**. The effects of pollutants (e.g., CS~2~) were analyzed per mile^2^.\
These data are a part of an on-going project with my Professor.  
This project is also available on my GitHub profile, rogue-economista under the test-repo repository. \newpage

```{=html}
<!-- 
Below is a numbered list of topics in the submission. Note that you can also add a table of contents to the YAML header -->
```

### A Rough Table of Contents

1.  First, a few papers on my topic, i.g., a Literature Review
2.  More on my Data

```{=html}
<!--
note that subtopics need to be indented. Below is an unordered list
-->
```

-   Cleaning Goals
    -   What worked from my Midterm plans
    -   What did not work from the Midterm and why
-   Analysis Goals
    -   Tables
    -   Viz

Here is a link to [more information on RMarkdown from RStudio](https://rmarkdown.rstudio.com/gallery.html)

This [link is the book](https://bookdown.org/yihui/rmarkdown-cookbook/) on RMarkdown written by the authors of knitr and other packages.

\newpage

### Background

Use this section to talk about what motivates your project. Is this furthering your learning in data science? Or are you using this data in an ongoing capacity?
Also, what has been written if anything about your topic? Give 1-2 citations.

### The Data

My data came from the Car Values data set. Here are the summary statistics of all variables.

```{r, import, include = FALSE}
car_values <- read_csv("data/raw/car_values.csv", show_col_types = FALSE)
```

```{r, sumstats-1}
summary(car_values)
```


\newpage

### A Multiple Regression Model

```{r, regression-1}

m <- lm(Price ~ Mileage + Doors + Cylinder, data = car_values)

```

The new object is a list of objects. Let's see what's inside!\
Alternatively, we can use inline R code to list the contents of m, e.g.,\
In this project I looked at the following **summary statistics** `r names(m)`. The mean car price was `r round(mean(car_values$Price), digits =2)`

```{r, regression-2}
names(m)

```

\newpage

Now let's take a look at the regression summary.

```{r, regression-3}
summary(m)
```

\newpage

```{r scatter-1, message=FALSE}
scatter <- ggplot(car_values, aes(x = Mileage, y = Price)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Car Data", subtitle = "this is the subtitle", caption = "use the caption to give more information about the viz above")

print(scatter)
```

### Next Steps

These are the Midterm goals that I was not (yet) able to accomplish and why.
